What is an Ancestral Recombination Graph?

### 2022-01-30 paper.tex
Some notes on the ARG

@@

\begin{itemize}
\item Terminology based around an inferred ancestry being a ``true''
ARG is impoverished and misleading. The only reasonable interpretation
of a ARG being true or not is whether it's a sample from the coalescent.
Most inference methods (and all that scale well) do not. In any case,
the coalescent is just one model, and a highly idealised one. It is
exceedingly unlikely that the actual ARG describing the ancestry
of any biological population is ``true'' ARG in this sense. The terminology is
impoverished because it reduces the possibilities available
to something either being an ARG or not, whereas there are many different
interesting properties of these structures that we should be able to
discuss and compare.
\end{itemize}

>>>>

\section*{Ancestral recombination graphs}
The coalescent~\citep{kingman1982coalescent,kingman1982genealogy,
hudson1983testing, tajima1983evolutionary}
provides the theoretical underpinning for much of contemporary population genetics.
Hudson extended the coalescent to include
recombination~\citep{hudson1983properties,hudson1990gene}
and the model has proved to be especially powerful basis for
simulations~\citep{hudson2002generating,kelleher2016efficient,kelleher2020coalescent,
baumdicker2021efficient}.
Hudson's algorithm is concerned with simulating the process of
the coalescent with recombination as efficiently as possible, and operates by
tracking segments of ancestral material carried by ancestors as
we go backwards in time. The details of Hudson's algorithm and how the
results of the ancestral process are represented are critical to later
discussions, and so we will explain the process in some detail.
See~\cite{kelleher2016efficient} for a more detailed
description of the algorithm, and how it may be implemented efficiently.

Hudson's algorithm operates by tracking the state of a set of ancestral
lineages as we go backwards in time. Each lineage consists of a list of
disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples the lineage is ancestral to over the interval.
(We also usually track the tree node associated with each segment, but
that is not important for our purposes here so we omit it.)
If we have $n$ samples and genome of length $L$, the process begins with $n$ lineages
of the form $[(0, L, 1)]$. The process then works backwards in time from
the present day as a series of random common ancestor or recombination events.
Recombination events occur at a rate determined by the amount and spatial
distribution of ancestral material among lineages.
Let $A$ be the set of ancestors at a given time $t$. Recombination events
happen at rate $\rho \nu / (L - 1)$ where
\[
\nu = \sum_{x \in A}\left( \max_{(\ell, r, a) \in x}r
    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
\]
is the number of available `links' that may be broken. Thus, the rate of
recombination is determined by the left and right-most extent of the
ancestral material carried by each lineage. At a recombination
event we choose one of these links uniformly and break it. Given a lineage
$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
$x_1$ and $x_2$ such that FILL IN DETAILS

When $k = |A|$ lineages are present, common ancestor events
occur at rate $k(k - 1) / 2$. In a common ancestor event two lineages
are chosed uniformly at random and their ancestry segments merged.
If we have overlapping intervals of ancestry from the two lineages,
say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
\emph{coalescence} occurs and ancestor represented by the current event
will be present as a node (at least) in the marginal trees covering
the interval $[\ell, r)$. The result of this coalescence is a segment
$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
ancesty for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
we have found the most recent common ancestor of all samples in
the interval $[\ell, r)$ and so we do not need to simulate its history any further.
Nonoverlapping intervals of ancestry from the two lineages are included
in the resulting lineage without changes. Eventually, as the process continues,
we find resultant lineages in which all segments have fully coalescenced,
and so the number of extant lineages gradually dwindles down to zero.

The Ancestral Recombination Graph (ARG) was later introduced
Griffiths~\citep{griffiths1991two,ethier1990two,griffiths1997ancestral}, and is a
closely related and complementary treatment of the same process. Where
Hudson's algorithm is focused on the efficient \emph{simulation} of the
process, the original ARG literature is focused on mathematical
results about the stochastic process. Griffiths and colleagues formulated the
ARG as a branching-coalescing stochastic process

NOTES for continuing:

\begin{itemize}
\item The ARG process does not track ancestral material. Recombination events
occur at rate $k \rho / 2$, and we don't complete until there's only one
lineage left. That means the process continues for an exponentially long time.
\item Discuss the way in which the re events work.
\item Discuss the rules for pulling out the trees. This is equivalent to doing
the interval logic above in Hudson's algorithm.
\end{itemize}

@@

\begin{itemize}
\item Terminology based around an inferred ancestry being a ``true''
ARG is impoverished and misleading. The only reasonable interpretation
of a ARG being true or not is whether it's a sample from the coalescent.
Most inference methods (and all that scale well) do not. In any case,
the coalescent is just one model, and a highly idealised one. It is
exceedingly unlikely that the actual ARG describing the ancestry
of any biological population is ``true'' ARG in this sense. The terminology is
impoverished because it reduces the possibilities available
to something either being an ARG or not, whereas there are many different
interesting properties of these structures that we should be able to
discuss and compare.
\item tskit can represent any ARG.
\end{itemize}
